---
title: "Day 2: Regression and ANOVA Exercise"
format: html
editor: source
---

```{r}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

```{r}
library(tidyverse)
```

## Part 1

```{r}
set.seed(100)
n <- 10000

# True Regression Coefficients
beta_0 <- 3
beta_1 <- 2
beta_2 <- 7.33
beta_3 <- 5
beta_4 <- 1.25
beta_5 <- 0
sigma2 <- 6

# Predictor Variables
x0 <- rep(1, n)
x1 <- rexp(n, rate = 2)
x2 <- rgeom(n, prob = 0.2)
x3 <- rnorm(n, mean = -2, sd = 5)
x4 <- rpois(n, 3)
x5 <- rbinom(n, 20, 0.5)

# Error Terms
epsilon <- rnorm(n, sd = sqrt(sigma2))
epsilon2 <- rnorm(n, sd = (2*x2 + 1))
epsilon3 <- rgamma(n, 2, 1)


y <- beta_0 + 
  (beta_1 * x1) + 
  (beta_2 * x2) + 
  (beta_3 * x3) +
  (beta_4 * x4) + 
  (beta_5 * x5) +
  epsilon

y2 <- beta_0 +
  (beta_1 * x1) + 
  (beta_2 * x2) +
  (beta_3 * x3) +
  (beta_4 * x4) +
  (beta_5 * x5) + 
  epsilon2

y3 <- beta_0 +
  (beta_1 * x1) +
  (beta_2 * x2) +
  (beta_3 * x3) +
  (beta_4 * x4) +
  (beta_5 * x5) +
  epsilon3

df <- tibble(x1, x2, x3, x4, x5, y)
df2 <- tibble(x1, x2, x3, x4, x5, y2)
df3 <- tibble(x1, x2, x3, x4, x5, y3)
```

First, run the code above to generate the three simulated datasets that we will be using for this exercise. Note that for every dataset, all the true regression coefficients used to generate the data are nonzero except for `beta_5`.

Fit a regression model on `df` using `x1`, `x2`, and `x3` as the predictor variables and run the appropriate regression diagnostics. Which OLS assumption is violated here? How do you think this affects our model estimates?

## Part 2

Fit a regression model on `df` using all possible predictor variables and run the appropriate regression diagnostics. Which OLS assumption is violated here? How do you think this affects our model estimates?

## Part 3

Fit a regression model on `df2` using all predictor variables except `x5`, and run the appropriate regression diagnostics. Which OLS assumption is violated? How does this affect our model estimates?

## Part 4

Fit a regression model on `df3` using all predictor variables except `x5`, and run the appropriate regression diagnostics. Which OLS assumption is violated? How does this affect our model estimates?
